### 2017-02-01 introduction.tex
Corrections to Abstract and Intro

Plasmas are fluid masses of charged particles, often formed when neutral atoms and molecules in a gas are ionized by external energy.  Instances of plasma include everything from florescent lightbulbs and flames to more exotic examples such as nuclear fusion reactor experiments, most stars, nebulas, and the interstellar medium.  In fact, about 99\% of visible matter in the known universe exists in the plasma state.  Of particular interest to this thesis is the plasma in Earth's ionosphere.

Ionospheric plasma is partially ionized, meaning only a fraction of the particles are ionized and the rest are neutral.  Considering the dynamics of plasmas therefore requires not only collisions between particles to be considered but also charge-charge interactions and external and internal electric and magnetic fields.  These dynamics can generate waves in the plasma, which cause density perturbations.  Similarly, variations in plasma temperature and electric and magnetic field strength can occur on a variety of scales.  This thesis focuses on mechanisms by which density variations in ionospheric plasma occur, often referred to as plasma structures.

\section{The Solar-Terrestrial Environment}
The solar-terrestrial environment begins with the Sun in the center of the solar system.  The Sun is composed of highly energized plasma that is gravitationally bound together.  The surface of the Sun is highly dynamic, populated with sunspots, which are dark and relatively cool regions with intense magnetic fields, as well as coronal holes, which are low-density areas characterized with a continuous outflow of plasma.  The intense magnetic fields associated with sunspots can break down releasing a large amount of energy and plasma, known as a solar flare.  Planet-sized masses of plasma known as coronal mass ejections (CMEs) can be ejected from the surface, which can be large enough to maintain an internal magnetic field as they travel through the solar system.  The frequency of sunspots and large outbursts of plasma changes with the 11-year-long solar cycle.  A large number of sunspots, high occurrence of fast and dense plasma outbursts, and a highly variable magnetic field indicate solar maximum.  Conversely, solar minimum is characterized by a low sunspot number and slow and relatively steady outflow.  This thesis examines factors that control irregularity production in the Earth's ionosphere, which can be strongly influenced by solar activity.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth,angle=0]{ste.pdf}
	\caption[Solar-terrestrial environment]{Diagram of the solar-terrestrial environment.  The solar wind, green arrows, contains both a steady stream of charged particles and the interplanetary magnetic field, dark blue.  The red region represents the magnetosheath, the shock layer formed due to the supersonic plasma outflow from the Sun.  The magnetopause encloses the magnetosphere dominated by the geomagnetic field and is shown by the black dashed line.  Closed magnetic field lines from the Earth are shown in grey, while open field lines are shown in light blue.  The ionosphere is shown in yellow above the Earth's surface.}
	\label{fig:ste}
\end{figure}

@@

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The focus of this body of work is the Earth's ionosphere, a region of the upper atmosphere that extends approximately between 50--1000 km in altitude where neutral gases have been partially ionized, resulting in free electrons and ions mixed with neutral particles.  Two competing processes occur in the ionosphere to create a peak in the electron density.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increases as altitude decreases because there are more neutral molecules available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination becomes a substantial factor, reducing the ion density.  The exact altitude where this density peak occurs varries with time of day, location, season, and solar cycle.  Representative model examples of electron density changes with altitude are shown in Figure \ref{fig:densprofile}, found from the International Reference Ionosphere (IRI) model run on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic (the IRI model is discussed further in Section \ref{sec:iri}).

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption[Ionosphere density profile]{Altitude profiles of electron density throughout the ionosphere found from the IRI model.  Typical extents of the \(D\), \(E\), and \(F\) regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

Because the concentrations of ions and neutral particles changes with altitude, each of these regions have distinct plasma physical properties that make them unique.  The main issue is whether the large-scale motion of a particle is controlled more by the magnetic field (the particle is magnetized) or by collisions with other particles (the particle is collisional).  This can be determined by comparing the collision frequency, \(\nu_\alpha\), to the gyrofrequency, \(\Omega_\alpha\), of a particular species, where \(\alpha\) refers to either electrons, \(e\), or ions, \(i\).  If \(\Omega_\alpha \gg \nu_\alpha\), the particle completes many gyrations between each collision, so its motion is determined mostly by the magnetic field and it is magnetized.  If \(\nu_\alpha \gg \Omega_\alpha\), the particle collides with other particles much more often than it complete a gyration, so it is considered collisional.  The gyrofrequency is dependent on the mass and charge of the particle and the strength of the magnetic field, \(\Omega_\alpha = q_\alpha B/m_\alpha\), so is roughly constant throughout the ionosphere because the magnetic field strength does not change much throughout these altitudes.  The collision frequency depends on the density of a species relative to the density of all other species in the plasma and can be calculated from a series of standard expressions \citep{Schunk1980,Schunk2009}.  In the \(E\) region, \(\Omega_e \gg \nu_e\), so electrons are magnetized.  However, the motion on ions is dominated by collisions between ions and neutrals (\(\nu_i > \Omega_i\)), so ions in the \(E\) region are considered collisional.  In the \(F\) region where the neutral density is much lower, both ions and electrons are magnetized, \(\Omega_\alpha \gg \nu_\alpha\).  These differences in how charged particles move impact the conductivity and plasma waves development, which is further discussed in Section \ref{sec:lit_instabilities}.

@@

Intermediate-scale plasma irregularities are those with sizes of 100 m -- 30 km \citep{Kelley2009}.  These are also referred to as scintillation-causing structures and are responsible for many of the aderse space weather effects that are observed on Earth.  One important example of a space weather effect is when the ionosphere changes the phase and amplitude of the original signal, called radio scintillation, which affects the ability of satellites to communicate with ground receivers.  This can cause communication blackouts or introduce large errors into GPS/GNSS (Global Positioning System/Global Navigation Satellite System) calculations, affecting navigation.

Small-scale structuring is generally considered to be all irregularities less than 100 m \citep{Kelley2009}.  Although small-scale irregularities are not responsible for scintillation that directly impact GPS/GNSS performance, they can be the easiest to study because of large datasets collected with a large number of ionospheric radars around the world that have been in operation for many decades (discussed further in Section \ref{sec:superdarn}).  Because electrons move far more easily along magnetic field lines than across them, density perturbations tend to be aligned with the magnetic field and are often known as field-aligned irregularities (FAI), Figure \ref{fig:polarirreg}d.  The irregularity classifications presented here are rough categories only and a high degree of coupling between different scales occurs, which makes plasma structures of all scales (large, intermediate, and small) important for plasma structuring processes.  For instance, large-scale structures such as polar patches may very well have internal intermediate- and/or small-scale structuring.  A turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs \citep{Kintner1985,Tsunoda1985}.  A description of some of the particular instruments and models that are used in this thesis follows.

@@

\subsection{Coherent Scatter Radars}
\label{sec:csr}
In this thesis, we make extensive use of Coherent Scatter Radars (CSR), which have been used to study the ionosphere for the last half century.  The radar transmits a radio wave, which can scatter off plasma structures in the ionosphere, and then be detected by the receiving antennas.  The time difference between when the signal was transmitted and when it was detected and the power and phase of the returned signal can then be used to determine characteristics of plasma irregularities.

@@

CSR systems can operate by transmitting pulses instead of continuously.  This allows the distance between the radar and the target to be found (based on the time difference between when the pulse is transmitted and received).  The size of the range gate is determined by the length of a radar pulse, Equation \ref{eqn:range_gate}
\begin{equation}
	\label{eqn:range_gate}
	\Delta r = \frac{c\Delta t}{2}.
\end{equation}
Here, \(\Delta r\) is the size of a range gate, \(\Delta t\) is the pulse length, and \(c\) is the speed of light.  Longer pulses result in larger range gates, resulting in poorer spatial resolution.

When the radar transmits continuously, backscatter will be received, but because it is impossible to know when that signal was transmitted, the difference in time between transmission and receiving cannot be used to calculate how far away the backscatter volume is.  Conversely, if the radar only transmits a pulse of a certain length of a certain length and then ``listens'' for its return, the time between transmission and return is known and the distance the pulse traveled can be calculated and therefore the location of the backscattering volume can be found \citep{Farley1972,Greenwald1983}.  However, using this simple single-pulse method another pulse cannot be transmitted until the first pulse returns, limiting the temporal resolution that can be achieved.  This can be improved by using a multipulse scheme, which will be discussed below \citep{Farley1972,Greenwald1983,Greenwald1985}.

A short time between pulse returns is necessary to allow Doppler velocities to be calculated, particularly at long ranges.  The maximum Doppler shifted frequency that can be observed by a radar is the Nyquist frequency, \(f_n\), which is equivalent to half the sampling frequency, \(f_s\), Equation \ref{eqn:nyquist}
\begin{equation}	
	\label{eqn:nyquist}
	f_n = \frac{f_s}{2}.
\end{equation}
Radars operating in single-pulse mode must have a very low sampling frequency, as described above, particularly at long ranges where the pulse must travel a long distance.  This severely limits the Doppler shift that can be measured, which places an upper limit on the LoS velocity that can be calculated with this technique.  For an ionospheric radar observing a structure 2000 km away, it would take about 13 ms for a signal traveling at the speed of light to travel to the structure and back to the radar, which corresponds to a sampling frequency of \(f_s = 75\) Hz.  The Nyquist frequency is then \(f_n = 37.5\) Hz, which corresponds to a maximum measurable Doppler velocity of \(\sim550\) m/s.  Flows in the ionosphere have been known to exceed 1000--2000 m/s, so single pulse sampling is insufficient for these purposes.

Instead of waiting for each individual signal to be received before transmitting the next, ionospheric radars typically employ a multipulse mode. In a multipulse mode, the radar transmits a series of pulses with different time intervals or lags between them.  This can introduce complications when backscatter from different pulses at different ranges is received by the radar at the same time, referred to as cross-range interference, but these effects can be mitigated using correlation techniques \citep{Farley1972}.  The smallest lag between two pulses is known as the multi-pulse increment, \(\tau\).  All other lags are integer multiples of \(\tau\), which allows the calculation of the corresponding lags of the complex autocorrelation function (ACF).  The ACF is used to find the spectral characteristics of the backscatter, from which the LoS velocity, spectral width, and backscatter power can be obtained.  Overall, multi-pulse techniques are generally considered far more appropriate for ionospheric studies than single pulse techniques \citep{Farley1972,Greenwald1983,Greenwald1985,Barthes1998,Ponomarenko2006}.

@@

Because convection maps require data covering as wide a range of magnetic local time sectors as possible, SuperDARN radars are continuously operational, usually in a common mode.  This creates a vast database of high-quality measurements of small-scale FAIs, which is very useful for other studies on irregularity occurrence and plasma structuring.  In the studies presented here, data were primarily used from the SuperDARN radars located at Rankin Inlet, Canada (RKN) and McMurdo Station, Antarctica (MCM).  These two radars have been shaded in Figure \ref{fig:superdarnmap}.  MCM is the only polar radar in the southern hemisphere that has been operational for more than 5 years.  Although the FoVs of other radars cover parts of the polar cap, these parts refer to either \(E\) region ranges or the furthest of the \(F\) region ranges, for which there is rarely much backscatter.  There are currently three polar radars in the northern polar cap, but the FoV of RKN overlaps the viewing area of an important complimentary instrument, the Resolute Bay Incoherent Scatter Radar FoV, discussed in Section \ref{sec:isr}, which makes it particularly useful for multi-instrument studies of the ionosphere.

@@

Analysis of the returned power spectrum reveals the electron density, electron and ion temperatures, and LoS ion velocity \citep{Evans1969,Rishbeth1985,Nicolls2007a}.  Using a model of chemical composition at different altitudes, the average ion mass can also be found.  In addition, 3D convection velocities and electric field vectors can be derived from LoS ion velocity measurements in some of the more modern ISRs by assuming both \(\vec{E}\) and \(\vec{V}_E\) are constant across the radar's FoV \citep{Heinselman2008}.  More recently, a new algorithm has been developed that used the Lagrange method of undetermined multipliers to generate 2D convection velocity and electric field vectors without making this limiting assumption \citep{Nicolls2014}.  Each instrument must be calibrated with known electron density measurements to account for noise and system constants \citep{Nicolls2007a}.  This is done using the plasma frequency measured from either the plasma line of the ISR spectra during summer daytime period or ionosonde measurements for all other times \citep{Bahcivan2010,Themens2014}.

@@

The first ISR was built in Arecibo, Puerto Rico in the 1950s \citep{Gordon1958}.  Since then, at least 8 other ISRs have been deployed around the world at equatorial, mid-, and high-latitudes.  A map of the ISR systems currently operational near the northern polar cap is shown in Figure \ref{fig:isrmap}.  The most recent advancement has been the development of Advanced Modular Incoherent Scatter Radars (AMISR) \citep{Heinselman2008}, of which there are currently three in use, one at the Poker Flat Rocket Range (PFISR), just north of Fairbanks, Alaska and two in Resolute Bay, Canada (RISR-N and RISR-C).  The advantage of these new systems is that they are electronically steerable, unlike older ISRs which consisted of a large dish that had to be physically moved to change the beam direction.  This allows beams to be transmitted in many different directions at a very high time cadence, giving the radar the capability to make measurements in multiple directions almost simultaneously \citep{Nicolls2007a,Nicolls2007b,Heinselman2008,Bahcivan2010}.  This is very important for creating 2D maps of ionospheric conditions and tracking how they change in time \citep{Semeter2009,Dahlgren2012a,Dahlgren2012b}.

AMISR systems typically have two types of pulses, a long pulse (LP) with 72-km range gates for \(F\)-region studies and an alternating code pulse (AC) with 4.5-km range gates gates for \(E\)-region studies.  The exact number and configuration of beams depends on the radar mode used and can be changed easily due to the system's electronic steering.  In the common WorldDay mode, there are typically 11 beams and data is collected at \(\sim\)1 min intervals, although this can change slightly between different renditions of the WorldDay mode.

One of the most important advantages of ISR systems over CSRs is the ability to directly measure electron density in the ionosphere.  However, because ISRs are much scarcer and are only typically run in WorldDay mode a few days out of every month, data are not as widely available.  In addition, ISRs do not actually observe small-scale coherent plasma structures like the SuperDARN HF network does.  However, they can directly measure large-scale structures and image density variations on scales larger that 100 km (limited by the number of beams and range gate size).  For these reasons, it is useful to use a combination of CSR and ISR measurements in ionospheric structuring studies, whenever possible.  The primary ISR used in the present volume of work is the north face of the AMISR system at Resolute Bay, Canada, RISR-N, shown in orange in Figure \ref{fig:isrmap}.  The FoV of RISR-N overlaps that of the RKN SuperDARN radar, red outline in Figure \ref{fig:isrmap}, making it ideal for these types of comparison studies in the polar cap.  The two experimental studies that employ both the SuperDARN RKN and RISR-N systems are presented in Chapters \ref{sec:paper1} and \ref{sec:paper3}.

@@

Although the IRI model is very useful for providing background plasma density conditions where no instruments are available to make measurements, it is important to bear in mind that it is a model and therefore only represents the average conditions.  Although it is fairly accurate at mid-latitudes \citep{Coisson2006,Bilitza2012}, it does not necessarily represent variations within the polar cap well \citep{Themens2014,Makarevich2015b}.  In particular, the height of the \(F\)-region peak tends to be underestimated in daytime and the bottomside thickness does not show realistic seasonal and diurnal variation \citep{Themens2014}.

Despite this, the IRI model is still a very important tool for establishing a density profile at any time in any location in the ionosphere.  An example of an altitude profile of plasma density provided by the IRI model was shown in Figure \ref{fig:densprofile}.  The displayed profiles shows how the electron density changes with altitude at local noon (day) and midnight (night) on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic.

One of the most important uses of the IRI model has been its integration into ray tracing simulations.  Throughout this work, it is valuable to determine the degree to which HF radar beams refract through a dense ionosphere.  This can be done through standard raytracing tools based on numerical solutions to the Hamiltonian raypath equations, as shown in Chapter \ref{sec:paper1} \citep{Haselgrove1963,Jones1975}, however these tools require a 2D density profile along the path that the radar beam travels.  It is possible to assume a Gaussian distribution or Chapman layer as a simple model of the F-region peak, but it is much more instructive to use output from the IRI model, as is done in this work, as it is sensitive to changes in latitude as well as seasonal and diurnal variations.

@@

\subsection{Polar Patch Formation}
\label{sec:lit_patches}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here the focus is on large regions of enhanced plasma density known as polar patches  \citep[e.g.][]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  One of the primary interests of this thesis is the relationships between irregularities of different scales and small-scale structuring.  This structuring has long been observed around polar patches \citep{Weber1984,Milan2002b,Moen2012}.  Although small-scale structures have also been observed around sun-aligned arcs \citep[e.g.][]{Koustov2012}, these structures also tend to be more complicated and often are surrounded by intense shears and electric fields \citep{Safargaleev2000,Aikio2002,Kozlovsky2007}.  Additionally, polar patches occur relatively frequently, providing a good opportunity to study irregularity coupling at different scales \citep{Rodger1996}.  In theory, many of the results presented in this work pertaining to the strong gradients on the edges of patches could also easily be applied to the edges of polar holes, but to date, far less attention has been paid to polar holes in the literature, making it challenging to accurately put observations in context \citep{Makarevich2015}.

>>>>

Plasmas are fluid masses of charged particles, often formed when neutral atoms and molecules in a gas are ionized by external energy.  Instances of plasma include everything from fluorescent lightbulbs and flames to more exotic examples such as nuclear fusion reactor experiments, most stars, nebulas, and the interstellar medium.  In fact, about 99\% of visible matter in the known universe exists in the plasma state.  Of particular interest to this thesis is the plasma in the Earth's ionosphere.

Ionospheric plasma is partially ionized, meaning that only a fraction of the particles are ionized and the rest are neutral.  Investigation of the dynamics of plasmas therefore requires not only collisions between particles to be considered but also charge-charge interactions and external and internal electric and magnetic fields.  These dynamics can generate waves in the plasma, which cause density perturbations.  Similarly, variations in plasma temperature and electric and magnetic field strength can occur on a variety of scales.  This thesis focuses on mechanisms by which density variations in ionospheric plasma occur, often referred to as plasma structures.

\section{The Solar-Terrestrial Environment}
The solar-terrestrial environment begins with the Sun in the center of the solar system.  The Sun is composed of highly energized plasma that is gravitationally bound together.  The surface of the Sun is highly dynamic, populated with sunspots, which are dark and relatively cool regions with intense magnetic fields, as well as coronal holes, which are low-density areas characterized with a continuous outflow of plasma.  The intense magnetic fields associated with sunspots can break down releasing a large amount of energy and plasma, known as a solar flare.  Planet-sized masses of plasma known as coronal mass ejections (CMEs) can be ejected from the surface, which can be large enough to maintain an internal magnetic field as they travel through the solar system.  The frequency of sunspots and large outbursts of plasma changes with the 11-year-long solar cycle.  A large number of sunspots, high occurrence of fast and dense plasma outbursts, and a highly variable magnetic field indicate solar maximum.  Conversely, solar minimum is characterized by a low sunspot number and slow and relatively steady outflow of plasma.  This thesis examines factors that control irregularity production in the Earth's ionosphere, which can be strongly influenced by solar activity.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth,angle=0]{ste.pdf}
	\caption[Solar-terrestrial environment]{Diagram of the solar-terrestrial environment.  The solar wind, green arrows, contains both a steady stream of charged particles and the interplanetary magnetic field, dark blue.  The red region represents the magnetosheath, the shock layer formed due to the supersonic plasma outflow from the Sun.  The magnetopause encloses the magnetosphere dominated by the geomagnetic field and is shown by the black dashed line.  Closed magnetic field lines from the Earth are shown in grey, while open field lines are shown in light blue.  The ionosphere is shown in yellow above the Earth's surface (not to scale).}
	\label{fig:ste}
\end{figure}

@@

\section{The Earth's Ionosphere}
\label{sec:ionosphere}
The focus of this body of work is the Earth's ionosphere, a region of the upper atmosphere that extends approximately 50--1000 km in altitude where neutral gases have been partially ionized, resulting in free electrons and ions mixed with neutral particles.  Two competing processes occur in the ionosphere to create a peak in the electron density.  Solar illumination and particle precipitation ionize neutrals, and the density of ions increases as altitude decreases because there are more neutral molecules available to ionize.  However, below a certain point, the concentration of ions becomes large enough that recombination becomes a substantial factor, reducing the ion density.  The exact altitude where this density peak occurs varries with time of day, location, season, and solar cycle.  Representative model examples of electron density changes with altitude are shown in Figure \ref{fig:densprofile}, found from the International Reference Ionosphere (IRI) model run on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic (the IRI model is discussed further in Section \ref{sec:iri}).

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption[Ionosphere density profiles]{Altitude profiles of electron density throughout the ionosphere found from the IRI model.  Typical extents of the \(D\), \(E\), and \(F\) regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

Because the concentrations of ions and neutral particles changes with altitude, each of these regions have distinct plasma physical properties that make them unique.  The main issue is whether the large-scale motion of a particle is controlled more by the magnetic field (the particle is magnetized) or by collisions with other particles (the particle is collisional).  This can be determined by comparing the collision frequency, \(\nu_\alpha\), to the gyrofrequency, \(\Omega_\alpha\), of a particular species, where \(\alpha\) refers to either electrons, \(e\), or ions, \(i\).  If \(\Omega_\alpha \gg \nu_\alpha\), the particle completes many gyrations between each collision, so its motion is determined mostly by the magnetic field and it is considered to be magnetized.  If \(\nu_\alpha \gg \Omega_\alpha\), the particle collides with other particles much more often than it completes a gyration, so it is considered collisional.  The gyrofrequency is dependent on the mass and charge of the particle and the strength of the magnetic field, \(\Omega_\alpha = q_\alpha B/m_\alpha\), so it is roughly constant throughout the ionosphere because the magnetic field strength does not change much throughout these altitudes.  The collision frequency depends on the density of a species relative to the density of all other species in the plasma and can be calculated from a series of standard expressions \citep{Schunk1980,Schunk2009}.  In the \(E\) region, \(\Omega_e \gg \nu_e\), so electrons are magnetized.  However, the motion of ions is dominated by collisions between ions and neutrals (\(\nu_i > \Omega_i\)), so ions in the \(E\) region are considered collisional.  In the \(F\) region where the neutral density is much lower, both ions and electrons are magnetized, \(\Omega_\alpha \gg \nu_\alpha\).  These differences in how charged particles move impact the conductivity and plasma waves development, which is further discussed in Section \ref{sec:lit_instabilities}.

@@

Intermediate-scale plasma irregularities are those with sizes of 100 m -- 30 km \citep{Kelley2009}.  These are also referred to as scintillation-causing structures and are responsible for many of the adverse space weather effects that are observed on Earth.  One important example of a space weather effect is when the ionosphere changes the phase and amplitude of the original radio signal, called radio scintillation, which affects the ability of satellites to communicate with ground receivers.  This can cause communication blackouts or introduce large errors into GPS/GNSS (Global Positioning System/Global Navigation Satellite System) calculations, affecting navigation.

Small-scale structuring is generally considered to be all irregularities less than 100 m \citep{Kelley2009}.  Although small-scale irregularities are not responsible for scintillation that directly impact GPS/GNSS performance, they can be the easiest to study because of extensive datasets collected with a large number of ionospheric radars around the world that have been in operation for many decades (discussed further in Section \ref{sec:superdarn}).  Because electrons move far more easily along magnetic field lines than across them, density perturbations tend to be aligned with the magnetic field and are often known as field-aligned irregularities (FAI), Figure \ref{fig:polarirreg}d.  The irregularity classifications presented here are rough categories only and a high degree of coupling occurs between different scales, which makes plasma structures of all scales (large, intermediate, and small) important for plasma structuring processes.  For instance, large-scale structures such as polar patches may very well have internal intermediate- and/or small-scale structuring.  A turbulent cascade is often assumed to connect these different scales, but because various instruments or techniques usually only observe structuring on a particular scale, it is difficult to find direct evidence of this or how exactly it occurs \citep{Kintner1985,Tsunoda1985}.  A description of some of the particular instruments and models that are used in this thesis follows.

@@

\subsection{Coherent Scatter Radars}
\label{sec:csr}
In this thesis, we make extensive use of coherent scatter radars (CSRs), which have been used to study the ionosphere for the last half century.  The radar transmits a radio wave, which can scatter off plasma structures in the ionosphere, and then be detected by the receiving antennas.  The time difference between when the signal was transmitted and when it was detected and the power and phase of the returned signal can then be used to determine characteristics of plasma irregularities.

@@

CSR systems can operate by transmitting pulses instead of continuously.  This allows the distance between the radar and the target to be found (based on the time difference between when the pulse is transmitted and received).  The size of the range gate is determined by the length of a radar pulse,
\begin{equation}
	\label{eqn:range_gate}
	\Delta r = \frac{c\Delta t}{2}.
\end{equation}
Here, \(\Delta r\) is the size of a range gate, \(\Delta t\) is the pulse length, and \(c\) is the speed of light.  Longer pulses result in larger range gates, resulting in poorer spatial resolution.

When the radar transmits continuously, backscatter will be received, but because it is impossible to know when that signal was transmitted, the difference in time between transmission and receiving cannot be used to calculate how far away the backscatter volume is.  Conversely, if the radar only transmits a pulse of a certain length of a certain length and then ``listens'' for its return, the time between transmission and return is known and the distance the pulse traveled can be calculated and therefore the location of the backscattering volume can be found \citep{Farley1972,Greenwald1983}.  However, using this simple single-pulse method, another pulse cannot be transmitted until the first pulse returns, limiting the temporal resolution that can be achieved.  This can be improved by using a multi-pulse scheme, which will be discussed below \citep{Farley1972,Greenwald1983,Greenwald1985}.

A short time between pulse returns is necessary to allow Doppler velocities to be calculated, particularly at long ranges.  The maximum Doppler shifted frequency that can be observed by a radar is the Nyquist frequency, \(f_n\), which is equivalent to half the sampling frequency, \(f_s\),
\begin{equation}	
	\label{eqn:nyquist}
	f_n = \frac{f_s}{2}.
\end{equation}
Radars operating in a single-pulse mode must have a very low sampling frequency, as described above, particularly at long ranges where the pulse must travel a long distance.  This severely limits the Doppler shift that can be measured, which places an upper limit on the LoS velocity that can be calculated with this technique.  For an ionospheric radar observing a structure 2000 km away, it would take about 13 ms for a signal traveling at the speed of light to travel to the structure and back to the radar, which corresponds to a sampling frequency of \(f_s = 75\) Hz.  The Nyquist frequency is then \(f_n = 37.5\) Hz, which corresponds to a maximum measurable Doppler velocity of \(\sim550\) m/s.  Flows in the ionosphere have been known to exceed 1000--2000 m/s, so single-pulse sampling is insufficient for these purposes.

Instead of waiting for each individual signal to be received before transmitting the next, ionospheric radars typically employ a multi-pulse mode. In a multi-pulse mode, the radar transmits a series of pulses with different time intervals or lags between them.  This can introduce complications when backscatter from different pulses at different ranges is received by the radar at the same time, referred to as cross-range interference, but these effects can be mitigated using correlation techniques \citep{Farley1972}.  The smallest lag between two pulses is known as the multi-pulse increment, \(\tau\).  All other lags are integer multiples of \(\tau\), which allows the calculation of the corresponding lags of the complex autocorrelation function (ACF).  The ACF is used to find the spectral characteristics of the backscatter, from which the LoS velocity, spectral width, and backscatter power can be obtained.  Overall, multi-pulse techniques are generally considered far more appropriate for ionospheric studies than single-pulse techniques \citep{Farley1972,Greenwald1983,Greenwald1985,Barthes1998,Ponomarenko2006}.

@@

Because convection maps require data covering as wide a range of magnetic local time sectors as possible, SuperDARN radars are continuously operational, usually in a common mode.  This creates a vast database of high-quality measurements of small-scale FAIs, which is very useful for other studies on irregularity occurrence and plasma structuring.  In the studies presented here, data were primarily used from the SuperDARN radars located at Rankin Inlet, Canada (RKN) and McMurdo Station, Antarctica (MCM).  These two radars have been shaded in Figure \ref{fig:superdarnmap}.  MCM is the only polar radar in the southern hemisphere that has been operational for more than 5 years.  Although the FoVs of other radars cover parts of the polar cap, these parts refer to either \(E\) region ranges or the furthest of the \(F\) region ranges, for which there is rarely much backscatter.  There are currently three polar radars in the northern polar cap, but the FoV of RKN overlaps the viewing area of an important complementary instrument, the Resolute Bay Incoherent Scatter Radar, discussed in Section \ref{sec:isr}, which makes it particularly useful for multi-instrument studies of the ionosphere.

@@

Analysis of the returned power spectrum reveals the electron density, electron and ion temperatures, and LoS ion velocity \citep{Evans1969,Rishbeth1985,Nicolls2007a}.  Using a model of chemical composition at different altitudes, the average ion mass can also be found.  In addition, 3D convection velocities and electric field vectors can be derived from LoS ion velocity measurements in some of the more modern ISRs by assuming that both \(\vec{E}\) and \(\vec{V}_E\) are constant across the radar's FoV \citep{Heinselman2008}.  More recently, a new algorithm has been developed that used the Lagrange method of undetermined multipliers to generate 2D convection velocity and electric field vectors without making this limiting assumption \citep{Nicolls2014}.  Each instrument must be calibrated with known electron density measurements to account for noise and system constants \citep{Nicolls2007a}.  This is done using the plasma frequency measured from either the plasma line of the ISR spectra during summer daytime period or ionosonde measurements for all other times \citep{Bahcivan2010,Themens2014}.

@@

The first ISR was built in Arecibo, Puerto Rico in the 1950s \citep{Gordon1958}.  Since then, at least 8 other ISRs have been deployed around the world at equatorial, mid-, and high-latitudes.  A map of the ISR systems currently operational near the northern polar cap is shown in Figure \ref{fig:isrmap}.  The most recent advancement has been the development of Advanced Modular Incoherent Scatter Radar (AMISR) systems \citep{Heinselman2008}, of which there are currently three in use, one at the Poker Flat Rocket Range (PFISR), just north of Fairbanks, Alaska and two in Resolute Bay, Canada (RISR-N and RISR-C).  The advantage of these new systems is that they are electronically steerable, unlike older ISRs which consisted of a large dish that had to be physically moved to change the beam direction.  This allows beams to be transmitted in many different directions at a very high time cadence, giving the radar the capability to make measurements in multiple directions almost simultaneously \citep{Nicolls2007a,Nicolls2007b,Heinselman2008,Bahcivan2010}.  This is very important for creating 2D maps of ionospheric conditions and tracking how they change in time \citep{Semeter2009,Dahlgren2012a,Dahlgren2012b}.

AMISR systems typically have two types of pulses, a long pulse (LP) with 72-km range gates for \(F\)-region studies and an alternating code pulse (AC) with 4.5-km range gates for \(E\)-region studies.  The exact number and configuration of beams depends on the radar mode used and can be changed easily due to the system's electronic steering.  In the common WorldDay mode, there are typically 11 beams and data is collected at \(\sim\)1 min intervals, although this can change slightly between different renditions of the WorldDay mode.

One of the most important advantages of ISR systems over CSRs is the ability to directly measure electron density in the ionosphere.  However, because ISRs are much scarcer and most are only typically run a few days out of every month, data are not as widely available.  In addition, ISRs do not actually observe small-scale coherent plasma structures like the SuperDARN array does.  However, they can directly measure large-scale structures and image density variations on scales larger that 100 km (limited by the number of beams and range gate size).  For these reasons, it is useful to use a combination of CSR and ISR measurements in ionospheric structuring studies, whenever possible.  The primary ISR used in the present volume of work is the north face of the AMISR system at Resolute Bay, Canada, RISR-N, shown in orange in Figure \ref{fig:isrmap}.  The FoV of RISR-N overlaps that of the SuperDARN RKN radar, red outline in Figure \ref{fig:isrmap}, making it ideal for these types of comparison studies in the polar cap.  The two experimental studies that employ both the SuperDARN RKN and RISR-N systems are presented in Chapters \ref{sec:paper1} and \ref{sec:paper3}.

@@

Although the IRI model is very useful for providing background plasma density conditions where no instruments are available to make measurements, it is important to bear in mind that it is a model and therefore only represents the average conditions.  Although it is fairly accurate at mid-latitudes \citep{Coisson2006,Bilitza2012}, it does not necessarily represent variations within the polar cap well \citep{Themens2014,Makarevich2015b}.  In particular, the height of the \(F\)-region peak tends to be underestimated in daytime and the bottomside thickness does not show realistic seasonal and diurnal variations \citep{Themens2014}.

Despite this, the IRI model is still a very important tool for establishing a density profile at any time in any location in the ionosphere.  An example of an altitude profile of plasma density provided by the IRI model was shown in Figure \ref{fig:densprofile}.  The displayed profiles show how the electron density changes with altitude at local noon (marked Day in Figure \ref{fig:densprofile}) and midnight (Night) on January 1, 2007 at \(75\deg\) N, \(0\deg\) E, geographic.

One of the most important uses of the IRI model has been its integration into ray tracing simulations.  Throughout this work, it is valuable to determine the degree to which HF radar beams refract through a dense ionosphere.  This can be done through standard raytracing tools based on numerical solutions to the Hamiltonian raypath equations, as shown in Chapter \ref{sec:paper1} \citep{Haselgrove1963,Jones1975}, however these tools require a 2D density profile along the path that the radar beam travels.  It is possible to assume a Gaussian distribution or Chapman layer as a simple model of the \(F\)-region peak, but it is much more instructive to use output from the IRI model, as is done in this work, as it is sensitive to changes in latitude as well as seasonal and diurnal variations.

@@

\subsection{Polar Patch Formation}
\label{sec:lit_patches}
Large-scale structuring in the polar cap can take a variety of forms, Section \ref{sec:polar_structure}, but here the focus is on large regions of enhanced plasma density known as polar patches  \citep[e.g.][]{Weber1984,Weber1986,Buchau1983,Buchau1985}.  One of the primary interests of this thesis is the relationships between large-scale plasma structures ans small-scale irregularity formation.  This structuring has long been observed around polar patches \citep{Weber1984,Milan2002b,Moen2012}.  Although small-scale structures have also been observed around sun-aligned arcs \citep[e.g.][]{Koustov2012}, these structures also tend to be more complicated and often are surrounded by intense shears and electric fields \citep{Safargaleev2000,Aikio2002,Kozlovsky2007}.  Additionally, polar patches occur relatively frequently, providing a good opportunity to study irregularity coupling at different scales \citep{Rodger1996}.  In theory, many of the results presented in this work pertaining to the strong gradients on the edges of patches could also easily be applied to the edges of polar holes, but to date, far less attention has been paid to polar holes in the literature, making it challenging to accurately put observations in context \citep{Makarevich2015}.
